### 2024-02-25 presentation/parts/presentation_hierarchical.tex
WIP

\section{Out-of-distribution detection}
% \subsection{Out-of-distribution detection}

\frame{
    \frametitle{Defining OOD detection}
    \begin{columns}

    \begin{column}{0.5\textwidth}

    Out-of-distribution (OOD) detection is about enabling models to distinguish the training data distribution $p(\xb)$ from any other distribution $\tilde{p}(\xb)$.

    We are concerned with doing this on a per-observation basis, i.e. answering the question:

    \begin{center}
        "Was $\xb$ sampled from $p(\xb)$ or not?"
    \end{center}

    % It is related to \textbf{anomaly detection} and \textbf{adversarial example detection}.
    \end{column}

    \begin{column}{0.5\textwidth}

    \begin{figure}[\textwidth]
        \centering
        \begin{tikzpicture}
        \begin{axis}[
          domain=0:13,samples=100,
          height=4.5cm, width=8cm,
          xtick=\empty,
          every axis plot/.append style={very thick,smooth}
        ]

        \addplot [name path=p1, color=red!50!black] {gauss(5,1)};
        \addplot [name path=p2, color=black] {gauss(8,1)};
        \path[name path=axis] (axis cs:0,0) -- (axis cs:10,0);
        \addlegendentry{$p(\xb)$}
        \addlegendentry{$\tilde{p}(\xb)$}

        \end{axis}
        \end{tikzpicture}
    \end{figure}
    \begin{figure}[\textwidth]
        \centering
        \begin{tikzpicture}
        \begin{axis}[
          domain=0:13,samples=100,
          height=4.5cm, width=8cm,
          every axis plot/.append style={very thick,smooth}
        ]

        \addplot [name path=p1, color=red!50!black] {gauss(4,0.5)};
        \addplot [name path=p2, color=black] {gauss(9,0.5)};
        \path[name path=axis] (axis cs:0,0) -- (axis cs:10,0);
        \addlegendentry{$p(\xb)$}
        \addlegendentry{$\tilde{p}(\xb)$}

        \end{axis}
        \end{tikzpicture}
    \end{figure}
    \end{column}

    \end{columns}
}

@@

\section{Latent variable models}

@@

\section{Identifying the issue}

@@

\begin{frame}
    \frametitle{Results on CIFAR10/SVHN}
    \begin{table}
        \centering
        \resizebox*{!}{0.87\textheight}{%
        \begin{tabular}{lrrr}
            \toprule
             Method & AUROC$\uparrow$ & AUPRC$\uparrow$ & FPR80$\downarrow$ \\
             \midrule
             \multicolumn{4}{c}{\textbf{CIFAR10 (in) / SVHN (out)}} \\
             \midrule
             \multicolumn{4}{l}{\textbf{Use prior knowledge of OOD}} \\
    Backgr. contrast. LR (PixelCNN) {\parencite{ren_likelihood_2019}}               & $0.930$ & $0.881$ & $0.066$ \\
    Backgr. contrast. LR (VAE) {\parencite{xiao_likelihood_2020}}                    & $0.265$ & - & - \\
    Outlier exposure {\parencite{hendrycks_deep_2019}}                              & $0.984$ & - & - \\
    Input complexity ($S$, Glow) \parencite{serra_input_2020}                   & $0.950$ & - & - \\
    Input complexity ($S$, PixelCNN++) \parencite{serra_input_2020}             & $0.929$ & - & - \\
    Input complexity ($S$, HVAE) (Ours) \parencite{serra_input_2020}\textsuperscript{\ref{footnote:serra_hierarchical_likelihood}} & $0.833$ & $0.855$ & $0.344$ \\
    \multicolumn{4}{l}{\textbf{Use in-distribution data labels $y$}} \\
    Mahalanobis distance {\parencite{lee_simple_2018}}                          & $0.991$ & - & -  \\
             \multicolumn{4}{l}{\textbf{No OOD-specific assumptions}} \\
             \multicolumn{4}{l}{\textit{- Ensembles}} \\
    WAIC, 5 models, Glow {\parencite{choi_waic_2019}}                          & $1.000$ & - & - \\
    WAIC, 5 models, PixelCNN {\parencite{ren_likelihood_2019}}                      & $0.628$ & $0.616$ & $0.657$ \\
             \multicolumn{4}{l}{\textit{- Not ensembles}} \\
    Likelihood regret \parencite{xiao_likelihood_2020}                               & $0.875$ & - & - \\
    $LLR^{>2}$ + HVAE (ours)                            & $0.811$ & $0.837$ & $0.394$ \\
    $LLR^{>2}$ + BIVA (ours)                            & $\mathbf{0.891}$ & $\mathbf{0.875}$ & $\mathbf{0.172}$ \\
             \bottomrule
        \end{tabular}%
        }
    \end{table}
\end{frame}

\frame{
    \frametitle{Results on diverse datasets}
    \begin{columns}
    \begin{column}{0.5\textwidth}

        \begin{table}[t]
            \centering
            \resizebox{0.9\textwidth}{!}{%
            \begin{tabular}{llrrr}
                \toprule
                 OOD dataset & Metric & AUROC$\uparrow$ & AUPRC$\uparrow$ & FPR80$\downarrow$ \\
                 \midrule
                 \multicolumn{5}{c}{\textbf{Trained on CIFAR10}} \\
                 \midrule
        SVHN          &  $LLR^{>2}$  &  0.811  &  0.837  &  0.394  \\
        CIFAR10       &  $LLR^{>1}$  &  0.469  &  0.479  &  0.835  \\
                 \midrule
                 \multicolumn{5}{c}{\textbf{Trained on SVHN}} \\
                 \midrule
        CIFAR10            &  $LLR^{>1}$       &  $0.939$  &  $0.950$  &  $0.052$  \\
        SVHN               &  $LLR^{>1}$       &  $0.489$  &  $0.484$  &  $0.799$  \\
            \bottomrule
            \end{tabular}
            }
        \end{table}
    \end{column}
    \begin{column}{0.5\textwidth}
        \begin{table}[t]
            \centering
            \resizebox{0.9\textwidth}{!}{%
            \begin{tabular}{llrrr}
                \toprule
                 OOD dataset & Metric & AUROC$\uparrow$ & AUPRC$\uparrow$ & FPR80$\downarrow$ \\
                 \midrule
                 \multicolumn{5}{c}{\textbf{Trained on FashionMNIST}} \\
                 \midrule
        MNIST                    & $LLR^{>1}$             &  0.986  &  0.987  &  0.011 \\
        notMNIST                 &  $LLR^{>1}$            &  0.998  &  0.998  &  0.000 \\
        KMNIST                   &  $LLR^{>1}$            &  0.974  &  0.977  &  0.017 \\
        Omniglot28x28            &  $LLR^{>2}$            &  1.000  &  1.000  &  0.000 \\
        Omniglot28x28Inverted    &  $LLR^{>1}$            &  0.954  &  0.954  &  0.050 \\
        SmallNORB28x28           &  $LLR^{>2}$            &  0.999  &  0.999  &  0.002 \\
        SmallNORB28x28Inverted   &  $LLR^{>2}$            &  0.941  &  0.946  &  0.069 \\
        FashionMNIST             &  $LLR^{>1}$            &  0.488  &  0.496  &  0.811 \\
                 \midrule
                 \multicolumn{5}{c}{\textbf{Trained on MNIST}} \\
                 \midrule
        FashionMNIST                   &  $LLR^{>1}$  &  $0.999$  &  $0.999$  &  $0.000$ \\
        notMNIST                       &  $LLR^{>1}$  &  $1.000$  &  $0.999$  &  $0.000$ \\
        KMNIST                         &  $LLR^{>1}$  &  $0.999$  &  $0.999$  &  $0.000$ \\
        Omniglot28x28                  &  $LLR^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
        Omniglot28x28Inverted          &  $LLR^{>1}$  &  $0.944$  &  $0.953$  &  $0.057$ \\
        SmallNORB28x28                 &  $LLR^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
        SmallNORB28x28Inverted         &  $LLR^{>1}$  &  $0.985$  &  $0.987$  &  $0.000$ \\
        MNIST                          &  $LLR^{>2}$  &  $0.515$  &  $0.507$  &  $0.792$ \\
                 \bottomrule
            \end{tabular}
            }
            % \caption{Additional results for the HVAE model trained on FashionMNIST. All results computed with 1000 importance samples.}
            % \label{tab:additional-results-fashionmnist}
        \end{table}
    \end{column}

    \end{columns}
}

>>>>

\section{hierarchical vaes know what they don't know}

\frame{
    \frametitle{Defining OOD detection}
    \begin{columns}
        \begin{column}{0.5\textwidth}
            Enable models to distinguish the training data distribution $p(\xb)$ from any other distribution $\tilde{p}(\xb)$.
            \vspace{3mm}
            Do this for any given single observation, i.e. answer the question:
            \vspace{3mm}
            \begin{center}
                "Was $\xb$ sampled from $p(\xb)$ or not?"
            \end{center}
        \end{column}
        \begin{column}{0.5\textwidth}
            \begin{figure}[\textwidth]
                \centering
                \begin{tikzpicture}
                        \begin{axis}[
                            name=axistop,
                            domain=0:13,samples=100,
                            height=4.5cm, width=8cm,
                            xtick=\empty,
                            every axis plot/.append style={very thick,smooth}
                        ]
                            \addplot [name path=p1, color=red!50!black] {gauss(5,1)};
                            \addplot [name path=p2, color=black] {gauss(8,1)};
                            \path[name path=axis] (axis cs:0,0) -- (axis cs:10,0);
                            \addlegendentry{$p(\xb)$}
                            \addlegendentry{$\tilde{p}(\xb)$}
                        \end{axis}
                        \path (axistop.south) ++ (0,-0.4cm) coordinate (test);
                        \begin{axis}[
                            name=axisbottom,
                            anchor=north,
                            shift=(test),
                            domain=0:13,samples=100,
                            height=4.5cm, width=8cm,
                            every axis plot/.append style={very thick,smooth}
                        ]
                            \addplot [name path=p1, color=red!50!black] {gauss(4,0.5)};
                            \addplot [name path=p2, color=black] {gauss(9,0.5)};
                            \path[name path=axis] (axis cs:0,0) -- (axis cs:10,0);
                            \addlegendentry{$p(\xb)$}
                            \addlegendentry{$\tilde{p}(\xb)$}
                        \end{axis}
                \end{tikzpicture}
            \end{figure}
        \end{column}
    \end{columns}
}

@@

% \section{Latent variable models}

@@

\begin{frame}
    \frametitle{Out-of-distribution detection with hierarchical VAEs}
    \begin{columns}
        \begin{column}{0.4\textwidth}
            \begin{itemize}
                \item Generative models learn to approximate the \textbf{data distribution} $p(\xb)$.
                \item The likelihood of the model given a sample $\xb$ is a measure of how well the model \textbf{explains the data}.
                \item \textbf{Model likelihood} have long been thought of as useful for OOD detection \cite{bishop_novelty_1994}.
            \end{itemize}
        \end{column}
        \begin{column}{0.6\textwidth}
            \begin{overprint}
                \onslide<2>
                \begin{figure}
                    \centering
                    \includegraphics[width=\textwidth]{../graphics/paper_hierarchical/densities-FashionMNIST-test-MNIST-test-bpp-k-0_sohau.pdf}
                \end{figure}
            \end{overprint}
        \end{column}
    \end{columns}
\end{frame}

% \section{Identifying the issue}

@@

\begin{frame}
    \frametitle{Results on CIFAR10/SVHN}
    \begin{table}
        \centering
        \resizebox*{!}{0.87\textheight}{%
        \begin{tabular}{lrrr}
            \toprule
            Method & AUROC$\uparrow$ & AUPRC$\uparrow$ & FPR80$\downarrow$ \\
            \midrule
            \multicolumn{4}{c}{\textbf{CIFAR10 (in) / SVHN (out)}} \\
            \midrule
            \multicolumn{4}{l}{\textbf{Use prior knowledge of OOD}} \\
            Backgr. contrast. LR (PixelCNN) {\parencite{ren_likelihood_2019}}               & $0.930$ & $0.881$ & $0.066$ \\
            Backgr. contrast. LR (VAE) {\parencite{xiao_likelihood_2020}}                    & $0.265$ & - & - \\
            Outlier exposure {\parencite{hendrycks_deep_2019}}                              & $0.984$ & - & - \\
            Input complexity ($S$, Glow) \parencite{serra_input_2020}                   & $0.950$ & - & - \\
            Input complexity ($S$, PixelCNN++) \parencite{serra_input_2020}             & $0.929$ & - & - \\
            Input complexity ($S$, HVAE) (Ours) \parencite{serra_input_2020} & $0.833$ & $0.855$ & $0.344$ \\
            \multicolumn{4}{l}{\textbf{Use in-distribution data labels $y$}} \\
            Mahalanobis distance {\parencite{lee_simple_2018}}                          & $0.991$ & - & -  \\
            \multicolumn{4}{l}{\textbf{No OOD-specific assumptions}} \\
            \multicolumn{4}{l}{\textit{- Ensembles}} \\
            WAIC, 5 models, Glow {\parencite{choi_waic_2019}}                          & $1.000$ & - & - \\
            WAIC, 5 models, PixelCNN {\parencite{ren_likelihood_2019}}                      & $0.628$ & $0.616$ & $0.657$ \\
            \multicolumn{4}{l}{\textit{- Not ensembles}} \\
            Likelihood regret \parencite{xiao_likelihood_2020}                               & $0.875$ & - & - \\
            $LLR^{>2}$ + HVAE (ours)                            & $0.811$ & $0.837$ & $0.394$ \\
            $LLR^{>2}$ + BIVA (ours)                            & $\mathbf{0.891}$ & $\mathbf{0.875}$ & $\mathbf{0.172}$ \\
            \bottomrule
        \end{tabular}%
        }
    \end{table}
\end{frame}

\frame{
    \frametitle{Results on diverse datasets}
    \begin{columns}
        \begin{column}{0.5\textwidth}
            \begin{table}[t]
                \centering
                \resizebox{0.9\textwidth}{!}{%
                \begin{tabular}{llrrr}
                    \toprule
                    OOD dataset & Metric & AUROC$\uparrow$ & AUPRC$\uparrow$ & FPR80$\downarrow$ \\
                    \midrule
                    \multicolumn{5}{c}{\textbf{Trained on CIFAR10}} \\
                    \midrule
                    SVHN          &  $LLR^{>2}$  &  0.811  &  0.837  &  0.394  \\
                    CIFAR10       &  $LLR^{>1}$  &  0.469  &  0.479  &  0.835  \\
                    \midrule
                    \multicolumn{5}{c}{\textbf{Trained on SVHN}} \\
                    \midrule
                    CIFAR10            &  $LLR^{>1}$       &  $0.939$  &  $0.950$  &  $0.052$  \\
                    SVHN               &  $LLR^{>1}$       &  $0.489$  &  $0.484$  &  $0.799$  \\
                    \bottomrule
                \end{tabular}
                }
            \end{table}
        \end{column}
        \begin{column}{0.5\textwidth}
            \begin{table}[t]
                \centering
                \resizebox{0.9\textwidth}{!}{%
                \begin{tabular}{llrrr}
                    \toprule
                    OOD dataset & Metric & AUROC$\uparrow$ & AUPRC$\uparrow$ & FPR80$\downarrow$ \\
                    \midrule
                    \multicolumn{5}{c}{\textbf{Trained on FashionMNIST}} \\
                    \midrule
                    MNIST                    & $LLR^{>1}$             &  0.986  &  0.987  &  0.011 \\
                    notMNIST                 &  $LLR^{>1}$            &  0.998  &  0.998  &  0.000 \\
                    KMNIST                   &  $LLR^{>1}$            &  0.974  &  0.977  &  0.017 \\
                    Omniglot28x28            &  $LLR^{>2}$            &  1.000  &  1.000  &  0.000 \\
                    Omniglot28x28Inverted    &  $LLR^{>1}$            &  0.954  &  0.954  &  0.050 \\
                    SmallNORB28x28           &  $LLR^{>2}$            &  0.999  &  0.999  &  0.002 \\
                    SmallNORB28x28Inverted   &  $LLR^{>2}$            &  0.941  &  0.946  &  0.069 \\
                    FashionMNIST             &  $LLR^{>1}$            &  0.488  &  0.496  &  0.811 \\
                    \midrule
                    \multicolumn{5}{c}{\textbf{Trained on MNIST}} \\
                    \midrule
                    FashionMNIST                   &  $LLR^{>1}$  &  $0.999$  &  $0.999$  &  $0.000$ \\
                    notMNIST                       &  $LLR^{>1}$  &  $1.000$  &  $0.999$  &  $0.000$ \\
                    KMNIST                         &  $LLR^{>1}$  &  $0.999$  &  $0.999$  &  $0.000$ \\
                    Omniglot28x28                  &  $LLR^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
                    Omniglot28x28Inverted          &  $LLR^{>1}$  &  $0.944$  &  $0.953$  &  $0.057$ \\
                    SmallNORB28x28                 &  $LLR^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
                    SmallNORB28x28Inverted         &  $LLR^{>1}$  &  $0.985$  &  $0.987$  &  $0.000$ \\
                    MNIST                          &  $LLR^{>2}$  &  $0.515$  &  $0.507$  &  $0.792$ \\
                    \bottomrule
                \end{tabular}
                }
            \end{table}
        \end{column}
    \end{columns}
}
